Co-methylation

### 2015-04-22 Rmarkdown/comethylation.Rmd
Begin re-write of Co-methylation.

# TODOs

- Try out Matthew Stephen's adaptive hierarchical shrinkage method on the estimated log odds ratios (see [https://github.com/stephens999/ash](https://github.com/stephens999/ash)).

## Chapter overview

In this chapter I describe two measures of within-sample co-methylation, _within-fragment co-methylation_ and _correlation of $\beta$-values_. Using both these measures I explore how co-methylation is affected by the distance between methylation loci and the genomic context of the methylation loci. I apply these methods to 40 whole-genome bisulfite-sequencing samples to study the between-sample variability of co-methylation.

Throughout this chapter I focus on co-methylation of CpGs, although the methods described are also applicable to non-CpG methylation loci. I do not explore co-methylation of non-CpG methylation loci because these loci are rarely methylated in the samples that I am studying.

\FloatBarrier
\section{Correlations of $\beta$-values}\label{subsec:beta_cor}

@@

\FloatBarrier
\section{Within-fragment co-methylation}\label{subsec:wf_cometh}

@@

Here I describe a simple method based on analysing methylation patterns at 2-tuples. For simplicity, I will use CpG 2-tuples, but the method is applicable to any methylation type. This method is implemented in the `cometh` function that is part of the `MethylationTuples` software.

The aim of this analysis is to address the question of 'how are methylation states on individual DNA fragments correlated as a function of the distance between loci?'. There are two steps in this analysis:

>>>>

\begin{chapabstract}
I describe two measures of within-sample co-methylation, _within-fragment co-methylation_ and _correlation of $\beta$-values_. The exploration of within-fragment co-methylation leads to a study of methods for estimating meaningful summaries of dependence in sparse $2 \times 2 \times K$ contingency tables.

Using estimates of both within-fragment co-methylation and correlation of $\beta$-values, I explore how co-methylation is affected by the distance between methylation loci and the genomic context of the methylation loci. I apply these methods to 40 whole-genome bisulfite-sequencing samples to study the between-sample variability of co-methylation.

Throughout this chapter I focus on co-methylation of CpGs, although the methods described are also applicable to non-CpG methylation loci.
\end{chapabstract}

\section{Correlations of $\beta$-values}\label{subsec:beta_cor}

@@

\subsection{Summary}

__TODO: Summarise the results for correlation of $\beta$-values.__

A limitation to analysing co-methylation using correlations of $\beta$-values is that the analysis is based on correlations of measurements that are aggregates from many thousands of DNA molecules. We are not measuring co-methylation on the scale at which the biological process of DNA methylation acts, namely at the level of individual DNA molecules. As noted by \citet{Ball:2009kh}, "the clonal feature of Illumina sequencing [allows us] to investigate whether co-methylation occurs at the single-molecule level". Before doing so, I first explore the challenge of estimating meaningful measures of dependence in sparse $2 \times 2 \times K$ contingency tables. While perhaps not yet apparent, this will be essential for the analysis of within-fragment co-methylation.

\section{Estimating dependence in sparse $2 \times 2 \times K$ contingency tables}

Consider two binary random variables, $X$ and $Y$. Let $\pi_{ij}, i, j = 1, 2$ denote the probability that a subject has response $i$ for variable $X$ and response $j$ for variable $Y$. We wish to estimate the _odds ratio_, which is defined as $\psi = \frac{\pi_{11} \pi_{22}}{\pi_{12} \pi_{21}}$. There are many reasons why we might wish to estimate the odds ratio (__CITE__), but our interest here is __WHAT__.

Suppose we observe $(X, Y)$ for $n$ samples. Denote by $n_{ij}, i, j = 1, 2$ the number of subjects with response $i$ for variable $X$ and response $j$ for variable $Y$. We use the 'dot' notation to denote the sum over the index it replaces, e.g., $n_{1 \cdot} = n_{11} + n_{12}$. The general form of this $2 \times 2$ contingency table is shown in Table \ref{tab:2x2}.

\begin{table}[h]
\centering
\begin{tabular}{@{}cccc@{}}
\cmidrule(l){2-4}
 & \multicolumn{3}{c}{Column} \\
 \cmidrule(l){2-4}
\multirow{3}{*}{Row} & & & \\
 & $n_{11}$       & $n_{12}$       & $n_{1 \cdot}$ \\
 & $n_{21}$       & $n_{22}$       & $n_{2 \cdot}$ \\
 & $n_{\cdot 1}$  & $n_{\cdot 2}$  & $n_{\cdot \cdot} = n$ \\ \cmidrule(l){2-4}
\end{tabular}
\caption{Notation for a $2 \times 2$ contingency table.}
\label{tab:2x2}
\end{table}

The simplest estimator of the odds ratio is the unconditional maximum likelihood estimator, $\widehat{\psi_{U}} = \frac{n_{11} n_{22}}{n_{12} n_{21}}$.

- Go through the various estimators, their pros and cons.
   - Conditional maximum likelihood estimator
- Explain the advantage of the "0.5-adjusted" unconditional estimator.
- Move to $2 \times 2 \times K$ tables
  - Go through same comparison of estimators
- Move to sparseness
  - Asymptopia I and II
  - Go through same comparison of estimators
  - The homogeneity assumption

 ## Setting for studying within-fragment co-methylation

 - Asymptopia II with sparse tables
 - Homogeneity??? Don't/can't know.

 ## Simulation

-

## Summary

\section{Within-fragment co-methylation}\label{subsec:wf_cometh}

@@

Here I describe a simple method based on analysing methylation patterns at 2-tuples. For simplicity, I will use CpG 2-tuples, but the method is applicable to any methylation type. This method is implemented in the `cometh()` function that is part of the `MethylationTuples` software.

The aim of this analysis is to address the questions:

1. How dependent are methylation states at nearby methylation loci on the same DNA fragment?
2. What influences this dependence?

There are two steps in this analysis:
